Puck Predictions: Unraveling the NHL Game Forecasting Riddle

### 2024-04-12 Latex/draft.tex
typos and graph

@@

As mentioned in the methods section, we trained an XGBoost model for each team in the NHL. One main benefit of this
approach is the ability to leverage the {\tt .predict\_proba()} method of XGBoost.
Using this method, we generated 
probabilistic predictions at various stages of a game, allowing us to 
plot the evolving probability of each team winning throughout the match
(see Figure~\ref{fig:xgboost_res}). 
This approach provided insights into momentum shifts (when one team 
has an advantage due to them having more players on the ice due to 
a penalty by the opposing team), key moments (such as goals and 
penalties, and critical plays influencing game outcomes).

@@

In Figure~\ref{tab:game_accuracies} we note that each model increased in accuracy as the game progressed. We also notice that the MCMC simulation method appears to have the highest accuracy after the start of the game while XGBoost appears to be the least accurate. We infer that the XGBoost models were overfit, since they appeared to perform very well on games within the training set. Also, during experimentation with the models, we noticed that the MCMC simulator tended to produce higher probabilities for the away team winning, and the XGBoost model tended to give higher probabilities to the home team winning. The Bayesian Network was unbiased towards the home or away team. We trained the MCMC model on the flow of events in the game, so it appears that the model picked up on a more predictive flow in how away teams play. Though the MCMC model was generally the most accurate, it takes about 2 minutes to simulate 50 games and get a single prediction while XGBoost and the Bayesian Network can make a prediction in under two seconds.

@@

In this project, we used three different methods to predict the outcomes of NHL games. We used a Bayesian Network to model the conditional dependencies of key features, XGBoost to model the play-by-play dynamics of each team, and an MCMC simulator to simulate the flow of events in a hockey game. We found that the MCMC simulator was the most accurate, with the Bayesian Network coming in second, and XGBoost coming in last. 
The win percentages predicted are more for entertainment and educational purposes, and should not be used for gambling or performance purposes, as they do not have a incredibly high degree of accuracy. There are many variables in a hockey game that affect win probability, and given more time and more data we would love to model other factors such as momentum, historical success, coaching, etc. Our project was limited by the data
we had access to and computational resources, but nonetheless we were able to produce a model that can predict the outcome of NHL games with a reasonable degree of accuracy.

>>>>

The models were each trained on prior seasons, and tested on the most recent 2022-2023 season. This split was chosen because in practice our model would have access to historical data but not have access to the current season's data as it happens.

@@

For XGBoost, we leveraged the {\tt .predict\_proba()} method to predict win probabilities after the model was fit on the data.
Using this method, we generated 
probabilistic predictions at various stages of a game, allowing us to 
plot the evolving probability of each team winning throughout the match
(see Figure~\ref{fig:xgboost_res}). 
This approach allowed for more insights into momentum shifts (when one team 
has an advantage due to them having more players on the ice due to 
a penalty by the opposing team), and key moments (such as goals and 
penalties, and critical plays influencing game outcomes) than the Bayesian network, resulting in more variance in the win probability graph.

@@

In Table~\ref{tab:game_accuracies} we note that each model increased in accuracy as the game progressed. 
The Bayesian network and MCMC model were the most accurate, while the XGBoost model had lower accuracy.
We infer that the XGBoost models were overfit, as on the training data it would regularly achieve higher than 90\% accuracy at most points during the game. 
Also, during experimentation with the models, we noticed that the MCMC simulator tended to produce higher probabilities for the away team winning, 
and the XGBoost model tended to give higher probabilities to the home team winning. 
The Bayesian Network was unbiased towards the home or away team. 
We trained the MCMC model on the flow of events in the game, so it appears that the model picked up on a more predictive flow in how away teams play. 
Though the MCMC model was generally the most accurate, it takes about 2 minutes to simulate 50 games and get a single prediction while XGBoost and the Bayesian Network can make a prediction in under two seconds.

@@

In this project, we used three different methods to predict the outcomes of NHL games. We used a Bayesian Network to model the conditional dependencies of key features, XGBoost to model the play-by-play dynamics of each team, and an MCMC simulator to simulate the flow of events in a hockey game. 
We found that the Bayesian network and MCMC simulator were more accurate than XGBoost, and Bayesian network also was significantly faster than the MCMC simulator. Therefore we can conclude that the Bayesian network is the best performer for this project, implying that the conditional dependencies and potential causal relationships between features are important for predicting the outcome of hockey games.
The win percentages predicted are more for entertainment and educational purposes, and should not be used for gambling or performance purposes, as they do not have a incredibly high degree of accuracy. There are many variables in a hockey game that affect win probability, and given more time and more data we would love to model other factors such as momentum, historical success, coaching, etc. Our project was limited by the data
we had access to and computational resources, but nonetheless we were able to produce a model that can predict the outcome of NHL games with a reasonable degree of accuracy depending on the time of game remaining.
